```{r  options, include=FALSE}
opts_chunk$set(fig.path="figs/", fig.keep='high')
```



# incidence: a R package for incidence computation and visualization

#### Computing incidence using *OutbreakTools*

Here is a simple example of incidence computation using *OutbreakTools*.

Load the package and data, and extract dates:
```{r  loadshit}
library(OutbreakTools)
data(ToyOutbreak)

dates <- get.data(ToyOutbreak, "DateInfected")

class(dates)
head(dates, 15)
range(dates)
```

Then we compute daily incidence:
```{r  incid}
incid.day <- get.incidence(dates)
incid.day

plot(incid.day, type="s", main="Daily incidence")
```

Same, with time windows of 2 days:
```{r  incid2}
incid.2days <- get.incidence(dates, interval=2)
incid.2days

plot(incid.2days, type="s", main="2 days incidence")
```


#### Core code example
Core computations can be found in OutbreakTools:

https://github.com/thibautjombart/OutbreakTools/blob/master/R/get.incidence.R#l10:l58


#### Note on cutting dates
By convention, when computing incidence the left bound (ealier date) is included and the right (later date) bound excluded.
