---
title: "Customize plots of incidence"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Customise graphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=8,
  fig.height=5,
  fig.path="figs-custom/"
)
```


This vignette provides some tips for the most common customisations of graphics produced by
`plot.incidence`. Our graphics use *ggplot2*, which is a distinct graphical system from base
graphics. If you want advanced customisation of your incidence plots, we recommend following an
introduction to *ggplot2*.


<br>
<br>
# Example data: simulated Ebola outbreak

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the package
[*outbreaks*](http://github.com/reconhub/outbreaks): `ebola.sim.clean`.

First, we load the data:
```{r, data}
library(outbreaks)
library(ggplot2)
library(incidence)

onset <- ebola.sim.clean$linelist$date.of.onset
class(onset)
head(onset)
```

We compute the daily incidence:
```{r, incid1}
i <- incidence(onset, interval = 7)
i

i.sex <- incidence(onset, interval = 7, group =  ebola.sim.clean$linelist$gender)
i.sex

i.hosp <- incidence(onset, interval = 7, group =  ebola.sim.clean$linelist$hospital)
i.hosp

```




<br>
<br>
# The `plot.incidence` function

When calling `plot` on an *incidence* object, the function `plot.incidence` is implicitely used. To access its documentation, use `?plot.incidence`. In this section, we illustrate existing customizations.


## Default behaviour

By default, the function uses grey for single time series, and colors from the color palette `pal1` when incidence is computed by groups:
```{r, default}
plot(i)
plot(i.sex)
plot(i.hosp)
```

However, some of these defaults can be altered through the various arguments of the function:
```{r, args}
args(incidence:::plot.incidence)
```




## Changing colors

### The default palette

A color palette is a function which outputs a specified number of colors. By default, the color used in *incidence* is called `pal1`. It's behaviour is different from usual palettes, in the sense that the first 4 colours are not interpolated:
```{r, pal1,   fig.height = 8}
par(mfrow=c(3,1), mar=c(4,2,1,1))
barplot(1:2, col = pal1(2))
barplot(1:4, col = pal1(4))
barplot(1:20, col = pal1(20))
```

This pallette also has a light and a dark version:
```{r, pal2,   fig.height = 8}
par(mfrow=c(3,1))
barplot(1:20, col = pal1dark(20), main = "palette:  pal1dark")
barplot(1:20, col = pal1(20), main = "palette:  pal1")
barplot(1:20, col = pal1light(20), main = "palette:  pal1light")
```



### Using different palettes

Other color palettes can be provided via `col_pal`. Various palettes are part of the base R distribution, and many more are provided in additional packages. We provide a couple of examples:

```{r, palettes}
plot(i.hosp, col_pal = rainbow)
plot(i.sex, col_pal = cm.colors)
```



### Specifying colors manually

Colors can be  specified manually using the argument `color`; note that whenever incidence is computed by groups, the number of colors must match the number of groups, otherwise `color` is ignored.

#### Example 1: changing a single color
```{r, colors1}
plot(i, color = "darkred")
```

#### Example 2: changing several colors (note that naming colors is optional)
```{r, colors2}
plot(i.sex, color = c(m = "orange2", f = "purple3"))
```

#### Example 3: using color to highlight specific groups
```{r, colors3}
plot(i.hosp, 
     color = c("#ac3973", "#6666ff", "white", "white", "white", "white"))
```





<br>
<br>
# Useful *ggplot2* tweaks

Numerous tweaks for *ggplot2* are documented online. In the following, we merely provide a few useful tips in the context of *incidence*.



##  Changing dates on the *x*-axis

### Changing date format

By default, the dates indicated on the *x*-axis of an incidence plot may not have the suitable format.
The package *scales* can be used to change the way dates are labelled (see `?strptime` for possible formats):

```{r, scales1}
library(scales)
plot(i) +
   scale_x_date(labels = date_format("%d %b %Y"))
```

And for another example, with a subset of the data (first 50 days), using more detailed dates and rotating the annotations:
```{r, scales2}
plot(i[1:50]) +  
  scale_x_date(labels = date_format("%a %d %B %Y")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12))
```

Note that you can save customisations for later use:
```{r, scales3}
rotate.big <- theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12))
```


### Changing the grid

The last example above illustrates that it can be useful to have denser annotations of the *x*-axis, especially over short time periods. Here, we provide an example where we try to zoom on the peak of the epidemic, using the data by hospital:

```{r, grid1}
plot(i.hosp)
```

Let us look at the data 40 days before and after the 1st of October:
```{r, grid2}
period <- as.Date("2014-10-01") + c(-40, 40)
i.zoom <- subset(i.hosp, from=period[1], to = period[2])
detailed.x <- scale_x_date(labels = date_format("%a %d %B %Y"), 
                           date_breaks = "2 weeks", 
                           date_minor_breaks="week")

plot(i.zoom, border="black") + detailed.x + rotate.big
```



## Changing the legend

The previous plot has a fairly large legend which we may want to move around. Let us save the plot as a new object `p` and customize the legend:

```{r, legend1}
p <- plot(i.zoom, border="black") + detailed.x + rotate.big
p + theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12), 
          legend.position="top", legend.direction="horizontal", 
          legend.title = element_blank())
```

