---
title: "Details of the incidence class"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Incidence class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




```{r, options, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width=7, 
  fig.height=5 
)

```



This vignette details the structure of *incidence* objects, as produced by the `incidence` function.

<br>

# Structure of an *incidence* object.

We generate a toy dataset of dates to examine the content of *incidence* objects.
```{r, data}
library(incidence)
set.seed(1)
dat <- sample(1:50, 200, replace = TRUE, prob = 1 + exp(1:50 * 0.1))
sex <- sample(c("female", "male"), 200, replace = TRUE)
```

The incidence by 48h period is computed as:
```{r, i}
i <- incidence(dat, interval = 2)
i
plot(i)
```

We also compute incidence by gender:
```{r, sex}
i.sex <- incidence(dat, interval = 2, group = sex)
i.sex
plot(i.sex)
```


The object `i` is a `list` with the class *incidence*:

```{r, names}
class(i)
is.list(i)
names(i)
```

Items in `i` can be accessed using the same indexing as any lists, but it's
safer to use the accessors for each item:

```{r, access}
## use name
head(i$dates)

head(get_dates(i))
```

In the following sections, we examine each of the components of the object.


## `$dates`

The `$dates` component contains a vector for all the dates for which incidence
have been computed, in the format of the input dataset (e.g. `Date`, `numeric`,
`integer`).

```{r, dates1}
class(get_dates(i))
class(dat)

get_dates(i)
```

The dates correspond to the lower bounds of the time intervals used as bins for
the incidence. Bins always include the lower bound and exclude the upper bound.
In the example provided above, this means that the first bin counts events that
happened at day 5-6, the second bin counts events from 7-8, etc.

## `$counts`

The `$counts` component contains the actual incidence, i.e. counts of events
for the defined bins. It is a `matrix` of `integers` where rows correspond to
time intervals, with one column for each group for which incidence is computed
(a single, unamed column if no groups were provided). If groups were provided,
columns are named after the groups. We illustrate the difference comparing the
two objects `i` and `i.sex`:

```{r, counts1}
class(get_counts(i))
storage.mode(get_counts(i))

get_counts(i)
get_counts(i.sex)
```

Note that a `data.frame` containing *dates* and *counts* can be obtained using
`as.data.frame`:

```{r, as.data.frame}
## basic conversion
as.data.frame(i)
as.data.frame(i.sex)

## long format for ggplot2
as.data.frame(i.sex, long = TRUE)
```

Note that `incidence` has an argument called `na_as_group` which is `TRUE` by
default, which will pool all missing groups into a separate group, in which
case it will be a separate column in `$counts`.

## `$timespan`

The `$timespan` component stores the length of the time period covered by the
object:

```{r, timespan}
i$timespan
range(i$dates)
diff(range(i$dates)) + 1
```






## `$interval`

The `$interval` component contains the length of the time interval for the bins:
```{r, interval}
i$interval
diff(i$dates)
```





## `$n`

The `$n` component stores the total number of events in the data:
```{r, n}
i$n
```

Note that to obtain the number of cases by groups, one can use:
```{r, n2}
apply(i.sex$counts, 2, sum)
```





## `$isoweeks`

The `$isoweeks` component is optional, and used to store iso weeks whenever they have been used. Iso weeks are used by default when weekly incidence is computed from dates (see argument `iso_week` in `?incidence`).
```{r, isoweek}
library(outbreaks)
dat <- ebola_sim$linelist$date_of_onset
i.7 <- incidence(dat, 7L)
i.7
i.7$isoweeks
```

The correspondence between iso weeks and dates can be derived by comparing the respective items:
```{r, isoweek2}
head(with(i.7, cbind.data.frame(dates, isoweeks)))
```

Both dates and iso weeks are returned when converting an `incidence` object to `data.frame`:
```{r, isoweek3}
head(as.data.frame(i.7))
```


